---
title: "2026-05-06_hw05"
author: "Chris"
date: "2026-05-06"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
    theme: cosmo
    css: style.css
---

The project is on GitHub: https://github.com/ChrisW12372/2026-05-06_p8106-hw05.git  


```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(caret)
library(kernlab)
library(e1071)
library(gbm)

set.seed(202605)
```


# Problem One

```{r}
set.seed(202605)
df_auto <- read_csv("data/auto.csv") |>
  mutate(
    mpg_cat = factor(mpg_cat),
    origin = factor(origin)
  )

data_split <- initial_split(df_auto, prop = 0.7)
df_auto_train <- training(data_split)
df_auto_test <- testing(data_split)
```


## Part I

```{r}
set.seed(202605)

cost_grid <- exp(seq(-5, 5, len = 100))

linear_tune_svc <- tune.svm(
  mpg_cat ~ cylinders + displacement + horsepower + weight + 
    acceleration + year + origin,
  data = df_auto_train,
  kernel = "linear",
  cost = cost_grid,
  scale = TRUE,
  tunecontrol = tune.control(cross = 10)
)

```



```{r}
summary(linear_tune_svc)
plot(linear_tune_svc)

linear_tune_svc$best.parameters
```

```{r}
# get the optimal classifier
best_svc <- linear_tune_svc$best.model
```


```{r}
svc_train_pred <- predict(best_svc, newdata = df_auto_train)
svc_train_error <- mean(svc_train_pred != df_auto_train$mpg_cat)

svc_train_error
```


```{r}
svc_test_pred <- predict(best_svc, newdata = df_auto_test)
svc_test_error <- mean(svc_test_pred != df_auto_test$mpg_cat)

svc_test_error
```


```{r}
df_svc_result <- tibble(
  Model = "SVC",
  Best_Cost = linear_tune_svc$best.parameters$cost,
  Training_Error = svc_train_error,
  Test_Error = svc_test_error
)

df_svc_result
```


I fit a support vector classifier with a linear kernel using the `e1071` package. The cost parameter was selected by 10-fold cross-validation. The best selected cost was `3.535`. Using this model, the training error rate was `0.0803`, and the test error rate was `0.0678`.  


## Part II

```{r}
set.seed(202605)

cost_grid <- exp(seq(-5, 5, len = 30))
gamma_grid <- exp(seq(-10, -2, len = 30))

radial_tune_svm <- tune.svm(
  mpg_cat ~ cylinders + displacement + horsepower + weight +
    acceleration + year + origin,
  data = df_auto_train,
  kernel = "radial",
  cost = cost_grid,
  gamma = gamma_grid,
  scale = TRUE,
  tunecontrol = tune.control(cross = 10)
)

summary(radial_tune_svm)
plot(
  radial_tune_svm,
  transform.y = log,
  transform.x = log,
  color.palette = terrain.colors
)
```


```{r}
best_radial_svm <- radial_tune_svm$best.model
```


```{r}
radial_train_pred <- predict(best_radial_svm, newdata = df_auto_train)

radial_train_error <- mean(radial_train_pred != df_auto_train$mpg_cat)

radial_train_error
```


```{r}
radial_test_pred <- predict(best_radial_svm, newdata = df_auto_test)

radial_test_error <- mean(radial_test_pred != df_auto_test$mpg_cat)

radial_test_error
```


```{r}
df_radial_result <- tibble(
  Model = "Radial SVM",
  Best_Cost = radial_tune_svm$best.parameters$cost,
  Best_Gamma = radial_tune_svm$best.parameters$gamma,
  Training_Error = radial_train_error,
  Test_Error = radial_test_error
)

df_radial_result
```


Using 10-fold cross-validation, the best radial SVM selected `cost = 26.466` and `gamma = 0.0113`. The resulting training error rate was `7.66%`, and the test error rate was `6.78%`. Compared with the linear SVC, the radial SVM slightly reduced the training error, but it did not improve the test error rate.  


## Part III

```{r}
set.seed(202605)

ctrl <- trainControl(
  method = "cv",
  number = 10
)

gbm_grid <- expand.grid(
  n.trees = c(100, 200, 500, 1000, 2000, 5000),
  interaction.depth = 1:4,
  shrinkage = c(0.005, 0.01, 0.05),
  n.minobsinnode = c(10)
)
```


```{r}
set.seed(202605)

gbm_fit <- train(
  mpg_cat ~ cylinders + displacement + horsepower + weight +
    acceleration + year + origin,
  data = df_auto_train,
  method = "gbm",
  tuneGrid = gbm_grid,
  trControl = ctrl,
  metric = "Accuracy",
  verbose = FALSE
)

gbm_fit
gbm_fit$bestTune
plot(gbm_fit)
```


```{r}
gbm_train_pred <- predict(gbm_fit, newdata = df_auto_train)
gbm_test_pred <- predict(gbm_fit, newdata = df_auto_test)

gbm_train_error <- mean(gbm_train_pred != df_auto_train$mpg_cat)
gbm_test_error <- mean(gbm_test_pred != df_auto_test$mpg_cat)

df_gbm_result <- tibble(
  Model = "Boosting",
  Best_n_trees = gbm_fit$bestTune$n.trees,
  Best_Depth = gbm_fit$bestTune$interaction.depth,
  Best_Shrinkage = gbm_fit$bestTune$shrinkage,
  Best_Minobsinnode = gbm_fit$bestTune$n.minobsinnode,
  Training_Error = gbm_train_error,
  Test_Error = gbm_test_error
)

df_gbm_result
```


```{r}
df_model_compare <- tibble(
  Model = c("SVC", "Radial SVM", "Boosting"),
  Training_Error = c(
    svc_train_error,
    radial_train_error,
    gbm_train_error
  ),
  Test_Error = c(
    svc_test_error,
    radial_test_error,
    gbm_test_error
  )
)

df_model_compare
```

The performance of the three models was compared using the test error rate. The SVC and radial SVM both had a test error rate of `0.0678`, while the boosting model had a lower test error rate of `0.0424`. Therefore, boosting provided the best predictive performance on this dataset. Neither the linear SVC nor the radial kernel SVM outperformed boosting.  





















































